Pomace: a Grappa for Non-Volatile Memory

### 2012-12-11 doc/nvm2013/paper.tex
edits mostly to first section

\title{Pomace: a Solid-State Grappa}
\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Holt$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  Preston Briggs$^{\dagger}$,
  Luis Ceze$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu}

\maketitle
%\begin{abstract}
%\input{abstract}
%\end{abstract}

We would like to present our software system Grappa, designed to scale
up emerging irregular applications such as graph analytics on
commodity clusters, and ask this community: What would be required to
adapt Grappa to exploit SSD, and how would doing so benefit this
application space?

\section{What is Grappa?}
Grappa is a latency tolerant runtime system for off-the-shelf clusters
that targets irregular applications, such as graph analytics.  These
applications are notoriously difficult to scale-up due to
data-dependent execution and memory access behaviors. The resulting
irregularity in computational demands hinders efforts to balance load
and to maintain spatial and temporal locality.  Resulting
inefficiencies are exacerbated at scale, as wasted energy rises with
the number of idling nodes, undersized network packets, and cache
misses.  Consequently, few irregular applications today make good use
of high-performance computers.  Those that do use complex, application
specific techniques to overlap computation with communication and were
developed by expert programmers who are in short supply.

Yet there is increasing demand for irregular applications.  Evidence
is the growing number of publications, conferences, and companies
focused on graph benchmarks, semantic databases, machine learning, and
“big data” discovery.

Grappa aims to mitigate the inefficiencies plaguing development and
execution of these important applications.  Grappa facilitates
expression of fine-grained task parallelism at the application
programming level.  Programmers attend only to expressing the
parallelism in their program, not to scheduling it.  At execution
time, Grappa’s task scheduler dynamically maps these tasks to x86
processor cores using work stealing heuristics similar to those
described in the literature.  What distinguishes Grappa from other
such runtime systems is that tasks implicitly yield at every long
latency operation, resulting in a context switch to another task.  In
this way, processors remain well-utilized even in the presence of
frequent references to remote memory.  Grappa thus tolerates many
microseconds of latency by multiplexing thousands of tasks per core.
In addition, Grappa aggregates references into packets destined for
common remote node locations.  In this way, Grappa’s design provides
for up to approximately 100 million remote references per second per
node using today’s commodity network hardware.  While similar latency
tolerance and aggregation techniques can be manually coded into some
applications, Grappa relieves the programmer of this burden.  In
addition, Grappa provides an API for reader-writer semantics for the
purpose of manipulating a local copy of any contiguous portion of the
shared address space, enabling programmers to exploit spatial and
temporal locality while maintaining a consistent global view.  With
these features, we have demonstrated performance on commodity systems
for highly irregular applications approaching that of the Cray XMT, a
fully custom system.

\subsection{What is the potential we imagine?}

@@

Emerging non-volatile memory technologies offer hope for an
alternative.  We imagine placing the entire data set on SSD or PCM and
either using Grappa directly to tolerate the latency much in the same
way as it is used now to tolerate network latency or to stage blocks
of data into RAM, using Grappa to tolerate the latency incurred by the
transfers.  We imagine the energy per available byte being much lower.
We imagine recovery from faults with most of the data residing in
non-volatile memory being faster.

Prior work only partially explores this potentiality. GraphChi
represents a class of approaches that use windowing methods in the
irregular application space, structuring the data on secondary storage
to support filling RAM with multiple interrelated subblocks sufficient
for each of a series of phases of computation.  Such approaches are
algorithmically efficient only when each successive phase requires
access to the entire data set.  MAGT addresses this limitation
introducing support for asynchronous computation on graphs under the
assumption that RAM is sized to accommodate data structures
proportional to vertex-count, still leaving edges on non-volatile
storage.  In addition, both methods apply only to single node systems:
Grappa provides a global view of memory on multinode systems.

@@

There are two problems with this design. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second,

>>>>

\title{Pomace: a Grappa for Non-Volatile Memory}
\authorinfo{Jacob Nelson$^{\dagger}$,
  Brandon Holt$^{\dagger}$,
  Brandon Myers$^{\dagger}$,
  Preston Briggs$^{\dagger}$,
  Luis Ceze$^{\dagger}$,
  Simon Kahan$^{{\dagger \ddagger}}$,
  Mark Oskin$^{\dagger}$
}{\textdagger University of Washington, \textdaggerdbl Pacific Northwest National Laboratory}{
  \{nelson, bholt, bdmyers, preston, luisceze, skahan, oskin\}@cs.washington.edu}

\maketitle
\begin{abstract}
We revisit Grappa, our software platform that scales in-memory irregular
applications on commodity clusters.
We contemplate options for adapting Grappa to use non-volatile memory technologies.
\end{abstract}

\section{What is Grappa?}
Grappa is a latency tolerant runtime system for off-the-shelf clusters
that targets in-memory irregular applications, such as interactive graph analytics.  These
applications are notoriously difficult to scale-up due to
data-dependent execution and memory access behaviors. The resulting
irregularity in computational demands hinders efforts to balance load
and to maintain spatial and temporal locality.  Resulting
inefficiencies are exacerbated at scale, as wasted energy rises with
the number of idling nodes, undersized network packets, and cache
misses.  Consequently, few irregular applications today make good use
of high-performance computers.  Those that do, use complex, application-specific techniques to overlap computation with communication and are
developed by expert programmers, who are in short supply.

Yet, there is increasing demand for irregular applications. Evidence
is the growing number of publications, conferences, and companies
focused on graph benchmarks, semantic databases, machine learning, and
“big data” discovery. Grappa aims to mitigate the inefficiencies plaguing development and execution of these important applications. The system (Figure~\ref{fig:grappa}) has three main software components:

\begin{figure}[t]
\begin{center}
  \includegraphics[width=0.95\columnwidth]{../asplos13/figs/system-overview}
\begin{minipage}{0.95\columnwidth}
  \caption{\label{fig:grappa} Grappa system overview}
\end{minipage}
\vspace{-3ex}
\end{center}
\end{figure}

\begin{description}

\item [Tasking system.] Our tasking system supports lightweight
multithreading to tolerate communication latency and global distributed
workstealing (i.e., tasks can be stolen from any node in the system), which
provides automated load balancing.  Grappa thus tolerates many
microseconds of latency by multiplexing thousands of tasks per x86 core.

\item[Distributed shared memory.] Our DSM system provides support for
fine-grain access to data anywhere in the system. It supports synchronization
operations on global data, explicit local caching of any memory in the system,
and support for operations on remote data (delegating operations to home node).
By tight integration with the tasking system and the
communication layer, our DSM system offers high aggregate random
access bandwidth to remote data.

\item[Communication layer.] Modern commodity networks
support high bandwidth only for large messages and irregular applications
frequently issue small requests.   Our communication layer aggregates these small requests into larger messages to better exploit network bandwidth, targeting 100 million remote references per second per node.  Aggregation is largely invisible to the application programmer.

% software developer but helps to improve network performance when applications read and write only small pieces of data.

\end{description}

Grappa thus facilitates
expression of fine-grained task parallelism at the application
programming level, employing that concurrency as needed to sustain throughput in the presence of long latency operations and to better utilize commodity networks.  Programmers attend only to expressing the parallelism in their program, not to scheduling it.  While similar latency
tolerance and aggregation techniques can be manually coded into some
applications, Grappa relieves the programmer of this burden.  In
addition, Grappa provides an API for reader-writer semantics for the
purpose of manipulating a local copy of any contiguous portion of the
shared address space, enabling programmers to exploit spatial and
temporal locality while maintaining a consistent global view.  With
these features, we have demonstrated performance on commodity systems
for highly irregular applications approaching that of the Cray XMT, a
fully custom system.

\subsection{What's the problem?}

@@

Emerging non-volatile memory technologies offer hope for an
alternative.  We imagine placing the entire data set on SSD or PCM and
either using Grappa directly to tolerate the latency, much in the same
way as it is used now to tolerate network latency, or to stage blocks
of data into RAM, using Grappa to tolerate the latency incurred by the
transfers.  We imagine the energy per available byte being much lower.
We imagine recovery from faults with most of the data residing in
non-volatile memory being faster.

Prior work only partially explores this potentiality. GraphChi
represents a class of approaches that use windowing methods in the
irregular application space, structuring the data on secondary storage
to support filling RAM with multiple interrelated subblocks sufficient
for each of a series of phases of computation.  Such approaches are
algorithmically efficient only when each successive phase requires
access to the entire data set.  MAGT addresses this limitation by
introducing support for asynchronous computation on graphs under the
assumption that RAM is sized to accommodate data structures
proportional to vertex-count, still leaving edges on non-volatile
storage.  In addition, both methods apply only to single node systems:
Grappa provides a global view of memory on multinode systems.

@@

There are two problems with this design. First, while each node
requires fewer chips, the overall cluster performance goal, and thus
chip requirement, is the same. Second, low single-threaded performance is detrimental to performance during periods of diminished application concurrency.
